---
title: "Behavior of Cyclistic Members vs Casual Riders"
author: "Jung Huh"
date: "8/7/2021"
output:   
  html_document:
    toc: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Business Task

### Questions to answer
How do annual members and casual riders use Cyclistic bikes differently?

### Task Statement
Find differences in behaviors and annual members and casual riders and provide insights on how Cyclystic can convert the casual riders into annual members

Stakeholders: Marketing Analytics Team, Director of Marketing

***

## Data Sources

### Data Details
* Data Source: Company Cloud Database ([link](https://divvy-tripdata.s3.amazonaws.com/index.html))
  + Reliable? Yes
  + Original? Yes
  + Comprehensive? Yes
  + Current? Yes
  + Cited? Yes
  

* Currently, data is organized by each month.
* Data has been made available by this [license](https://www.divvybikes.com/data-license-agreement)

* Categorization of Member vs Casual
  + Single rides & Full day purchases are logged under Casual
  + Annual membership purcahses are logged under Member

***

## Cleaning and Preparing Data

### Importing Data

#### Install packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(data.table)
```
#### Collect data

```{r message=FALSE, warning=FALSE}
setwd("C:/Users/Albert/Desktop/Cyclistic_Data/CSV")

Jun_2021 <- read_csv("202106-divvy-tripdata.csv")
May_2021 <- read_csv("202105-divvy-tripdata.csv")
Apr_2021 <- read_csv("202104-divvy-tripdata.csv")
Mar_2021 <- read_csv("202103-divvy-tripdata.csv")
Feb_2021 <- read_csv("202102-divvy-tripdata.csv")
Jan_2021 <- read_csv("202101-divvy-tripdata.csv")
Dec_2020 <- read_csv("202012-divvy-tripdata.csv")
Nov_2020 <- read_csv("202011-divvy-tripdata.csv")
Oct_2020 <- read_csv("202010-divvy-tripdata.csv")
Sep_2020 <- read_csv("202009-divvy-tripdata.csv")
Aug_2020 <- read_csv("202008-divvy-tripdata.csv")
Jul_2020 <- read_csv("202007-divvy-tripdata.csv")
```
### Cleaning and Preparing Data
#### Inspecting column to combine data into a single dataframe

* Extracting column names from latest monthly data and comparing it to rest using check_columns function
* check_columns function
  + returns "Matching Columns' if column names match
  + returns "Mismatching Columns' if column names do **not** match


```{r message=FALSE, warning=FALSE}
print(colnames(Jun_2021))
column_names <- colnames(Jun_2021)

check_columns <- function(monthlydata){
  if (column_names != colnames(monthlydata)){
    print('Mismatching Columns')
  } else {
    print('Mathcing Columns')
  }
}

check_columns(May_2021)
check_columns(Apr_2021)
check_columns(Mar_2021)
check_columns(Feb_2021)
check_columns(Jan_2021)
check_columns(Dec_2020)
check_columns(Nov_2020)
check_columns(Oct_2020)
check_columns(Sep_2020)
check_columns(Aug_2020)
check_columns(Jul_2020)

```
#### Inspecting column attribute types

```{r}

str(Jun_2021)
str(May_2021)
str(Apr_2021)
str(Mar_2021)
str(Feb_2021)
str(Jan_2021)
str(Dec_2020)
str(Nov_2020)
str(Oct_2020)
str(Sep_2020)
str(Aug_2020)
str(Jul_2020)
```
* Found out that start_station_id and end_station_id are in different data types in Nov_2020 - Jun_2020

#### Stacking monthly data into a data frame

* Changing data types for congruency
* Creating a master data frame called oneYear_data

```{r message=FALSE, warning=FALSE}
Nov_2020 <- mutate(Nov_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))
Oct_2020 <- mutate(Oct_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))
Sep_2020 <- mutate(Sep_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))
Aug_2020 <- mutate(Aug_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))
Jul_2020 <- mutate(Jul_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))


oneYear_data <- bind_rows(Jun_2021, May_2021, Apr_2021, Mar_2021, Feb_2021, Jan_2021, Dec_2020, Nov_2020, Oct_2020, Sep_2020, Aug_2020, Jul_2020)
head(oneYear_data)
```


#### Checking for data consistency
* Checking for consistency in categorical columns

```{r}
table(oneYear_data$member_casual)
table(oneYear_data$rideable_type)
```

#### Inspecting the new data table

```{r}
colnames(oneYear_data)
nrow(oneYear_data)
dim(oneYear_data)
head(oneYear_data)
str(oneYear_data)
summary(oneYear_data)
```
#### Adding new columns

##### Seprating day, month, year from started_at

```{r}
oneYear_data$date <- as.Date(oneYear_data$started_at)
oneYear_data$month <- format(as.Date(oneYear_data$date), '%m')
oneYear_data$day <- format(as.Date(oneYear_data$date), '%d')
oneYear_data$year <- format(as.Date(oneYear_data$date), '%Y')
oneYear_data$day_of_week <- format(as.Date(oneYear_data$date), '%A')
```


##### Adding 'ride_duration' column

```{r}
oneYear_data$ride_duration <- difftime(oneYear_data$ended_at, oneYear_data$started_at)
```

#### Removing 'bad' data

* Removing data with negative or zero (less than 1 second) ride_duration
  + Some rows had negative duration 
  + After some research, we can conclude that these bikes were taken out of circulation for quality control
```{r}
negative_duration <- filter(oneYear_data, ride_duration < 1)
str(negative_duration)
nrow(negative_duration)
```
* Removing data with ride duration of one **day** and more
  + Upon reviewing company's policy on bike rentals of more than 24 hours, bikes that have been rented for 24+ hours are treated as **stolen** or **missing**
  
```{r}
oneYear_data_V3 <-filter(oneYear_data, ride_duration >= 1 & ride_duration < 60*60*24)
nrow(oneYear_data_V3)
```
* Removing data resulting from internal tests
  + Testing data are not relevant to the analysis

```{r}
nrow(subset(oneYear_data_V3, start_station_name %like% "TEST"))
nrow(subset(oneYear_data_V3, start_station_name %like% "test"))
nrow(subset(oneYear_data_V3, start_station_name %like% "Test"))
```
```{r}
oneYear_data_V4 <- filter(oneYear_data_V3, !(oneYear_data_V3$start_station_name %like% "TEST" | oneYear_data_V3$start_station_name %like% "test"))
nrow(oneYear_data_V4)
```


## Analysis

### Taking a look at some statistical values

#### Average ride duration of Members vs Casuals

```{r}
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual, FUN = mean)
```
#### Median ride duration of Members vs Casuals

```{r}
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual, FUN = median)
```

#### Max. and Min. ride duration of Members vc Casuals

```{r}
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual, FUN = max)
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual, FUN = min)
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual, FUN = sd)
```

#### Examining ride duration for a weekday for Members vs Casuals


```{r}
oneYear_data_V4$day_of_week <- ordered(oneYear_data_V4$day_of_week, levels = c("Sunday", "Monday", 'Tuesday', 'Wednesday', 'Thursday', 'Friday' , 'Saturday'))
aggregate(oneYear_data_V4$ride_duration ~ oneYear_data_V4$member_casual + oneYear_data_V4$day_of_week, FUN = mean)
```

#### Examining number of rides and ride duration for a weekday for Members vs Casuals


```{r message=FALSE, warning=FALSE}
oneYear_data_V4 %>% 
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
           group_by(member_casual, weekday) %>% 
           summarise(number_of_rides = n(), average_duration = mean(ride_duration)) %>% 
           arrange(member_casual, weekday)
```

## Data Viz and Key Fidings

### Members vs Casuals - Number of Rides each day of the Week

```{r}
oneYear_data_V4 %>% 
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
           group_by(member_casual, weekday) %>% 
           summarise(number_of_rides = n(), average_duration = mean(ride_duration)) %>% 
           arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + xlab("Day of the Week") + ylab('Number of Rides') + ggtitle("Number of Rides of Members vs Casuals")

ggsave('Number_of_Rides.png')
```


### Members vs Casuals - Average Ride Duration


```{r}
oneYear_data_V4 %>% 
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
           group_by(member_casual, weekday) %>% 
           summarise(number_of_rides = n(), average_duration = mean(ride_duration)) %>% 
           arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + xlab("Day of the Week") + ylab('Average Ride Duration (sec)') + ggtitle("Average Ride Durations of Members vs Casuals")

ggsave('Average_Ride_Duration.png')
```


### Members vs Casuals - Number of Rides each Month

```{r}
oneYear_data_V4 %>% 
  group_by(member_casual, month) %>% 
  summarise(total_rides = n()) %>% 
  ggplot(aes(x = month, y = total_rides, fill = member_casual)) +
  geom_col(position = 'dodge') + xlab("Months") +ylab("Number of Rides") +ggtitle("Number of Rides each Month")
```
### Members vs Casuals - Number of Rides by each Hour

```{r}
oneYear_data_V4$time <- format(oneYear_data_V4$started_at, format = '%H:%M')
oneYear_data_V4$time <- as.POSIXct(oneYear_data_V4$time, format = '%H')
oneYear_data_V4 %>% 
  group_by(member_casual, time) %>% 
  summarise(total_rides = n()) %>% 
  ggplot(aes(x = time, y = total_rides, color = member_casual, group = member_casual)) +
  geom_line() + scale_x_datetime(date_breaks = "1 hour", minor_breaks = NULL,
date_labels = "%H", expand = c(0,0)) + xlab('Hour') + ylab('Total Rides') +ggtitle("Number of Rides by Hour")
```

### Key Findings and Disclaimer

#### Key Findings

* Ride Numbers:
  + Casual riders' ride numbers increase significantly on weekends (Friday - Sunday)
  + **56%** of all casual riders trips were recorded on weekends
  + Member riders' ride numbers stay mostly consistent except beginning of the week (Monday & Tuesday)

```{r}
nrow(subset(oneYear_data_V4, member_casual == 'casual' & (day_of_week == 'Friday' | day_of_week == 'Saturday' | day_of_week == 'Sunday'))) / nrow(subset(oneYear_data_V4, member_casual == 'casual'))
```


* Ride Duration:
  + Casual riders' average ride duration increases significantly on weekends(Friday - Sunday)
  + Member riders' average ride duration stays consistent except on Saturdays & Sundays
  + In comparison, trips of casual riders were **twice** as long as trips of members
  
* Casual riders are more actively renting bikes in the warmer months
  + **48%** casual rider trips were recorded between June to August
  + Only 38% or member rider trips were recorded in the same time frame
  
```{r}
nrow(subset(oneYear_data_V4, member_casual == 'casual' & (month == "06" | month == "07" | month == "08"))) / nrow(subset(oneYear_data_V4, member_casual == 'casual'))
```
```{r}
nrow(subset(oneYear_data_V4, member_casual == 'member' & (month == "06" | month == "07" | month == "08"))) / nrow(subset(oneYear_data_V4, member_casual == 'member'))
```


* Casual riders tend to rent significantly less in the 'rush' hours
  + Referencing the plot "Number of Rides by Hour", the biggest differences are displayed at approximately between Hour 7, 8, 14, and 15.
  + Casual riders
  




#### Disclaimer 

* Data Removal:
  + Data with negative ride duration were removed 
  + Data with ride duration of greater than 24 hours were removed
  
* Possible Bias:
  + May have introduced sampling bias by removing rides where ride duration was negative or greater than 24 hours
  
## Recommendations

### First

* Create marketing strategy and new membership program targeting Casual riders on weekends (Friday - Sunday)
  + Casual riders ride **more and longer** on weekends
  + New membership program more suited for weekends riders is strong recommended

### Second

* Accommodate riders quicker in the morning rush hours
  + Biggest difference in number of rides between members and non-members are found during rush hours
  + Casual riders may opt out of riding a bike if it takes more than a minute or two to rent/find one
  + Consolidating the rental process during rush hours recommended  

### Third

* Create a database with unique ID for each rider instead of each trip
  + It will enable deeper analysis on specific subsets of riders
  + Understanding characteristics of casual riders and their trips can provide valuable information that can be used in user conversion
